---
title: "Q$"
author: "Gerald"
date: "3/4/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:/Users/geral/Repository/Coursework/Dataset/")
```

```{r}
setwd("C:/Users/geral/Repository/Coursework/Dataset")
getwd()
```

```{r}
library(DBI)
library(RSQLite)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyr)
library(tidyverse)
library(RColorBrewer)
library(car)
library(broom)
```

# Basic Setup
For this we will be taking a look at the year 2005, selecting specific periods and airports to observe any cascading effects.

### Connecting to DB Browser
```{r}
conn <- dbConnect(RSQLite::SQLite(), "dataset.db")
```

```{r}
airports <- dbGetQuery(conn, "
                       SELECT * 
                       FROM airports")
```

```{r}
carriers <- dbGetQuery(conn, "
                       SELECT * 
                       FROM carriers")
```

```{r}
Q4_2005 <- dbGetQuery(conn, "
                 SELECT *
                 FROM '2005'
                 WHERE ArrTime <= 2359 AND DepTime <=2359")

Q4_2005_narm <- na.omit(Q4_2005)
rm(Q4_2005)
str(Q4_2005_narm)
# Observations : 6,982,428 
```

### Extraction Functions
We create 4 functions to extract arrival and departure delays TO our target airport and arrival and departure delays FROM our target airport.

```{r}
# function to extract data depdelay TO a target airport
to_depdelay <- function(mon, day, dest){
  # Dest must be in 'dest'
  Q4_2005_narm %>% 
    filter(Month == mon, DayofMonth == day, Cancelled == 0, Diverted == 0,
           Dest == dest, DepTime <= 2359, ArrTime <= 2359) %>% 
    select(TailNum, CRSDepTime,DepDelay) %>% 
    arrange(CRSDepTime) %>% 
    rename('Scheduled_Time' = 'CRSDepTime',
           'Delay' = 'DepDelay')
}

# sample
# testtoATLDepDelay <- tp_depdelay(7,7, 'ATL')
# creating a category column 
# i <- nrows(testtoATLDepDelay)
# testtoATLDepDelay$cat <- rep('Departure to ATL', i)
```

```{r}
# function to extract data arrdelay TO a target airport
to_arrdelay <- function(mon, day, dest){
  # Dest must be in 'dest'
  Q4_2005_narm %>% 
    filter(Month == mon, DayofMonth == day, Cancelled == 0, Diverted == 0,
           Dest == dest, DepTime <= 2359, ArrTime <= 2359) %>% 
    select(TailNum, CRSArrTime,ArrDelay) %>% 
    arrange(CRSArrTime) %>% 
    rename('Scheduled_Time' = 'CRSArrTime',
           'Delay' = 'ArrDelay')
}

# sample
# testtoATLArrDelay <- tp_arrdelay(1,1, 'ATL')
# creating a category column
# i <- nrows(testtoATLArrDelay)
# testtoATLArrDelay$cat <- rep('Arrival at ATL', i)
```

```{r}
# function to extract data depdelay FROM a target airport
from_depdelay <- function(mon, day, ori){
  # Origin must be in 'ori'
  Q4_2005_narm %>% 
    filter(Month == mon, DayofMonth == day, Cancelled == 0, Diverted == 0,
           Origin == ori, DepTime <= 2359, ArrTime <= 2359) %>% 
    select(TailNum, CRSDepTime,DepDelay) %>% 
    arrange(CRSDepTime) %>% 
    rename('Scheduled_Time' = 'CRSDepTime',
           'Delay' = 'DepDelay')
}
```

```{r}
# function to extract data arrdelay FROM a target airport
from_arrdelay <- function(mon, day, ori){
  # Origin must be in 'ori'
  Q4_2005_narm %>% 
    filter(Month == mon, DayofMonth == day, Cancelled == 0, Diverted == 0,
           Origin == ori, DepTime <= 2359, ArrTime <= 2359) %>% 
    select(TailNum, CRSArrTime,ArrDelay) %>% 
    arrange(CRSArrTime) %>% 
    rename('Scheduled_Time' = 'CRSArrTime',
           'Delay' = 'ArrDelay')  
}
```

## Checks
We will be observing airports which have the highest air traffic.
```{r}
count <- Q4_2005_narm %>% 
  count(Dest) %>% 
  arrange(n)
tail(count)

# Destination with the most flights is: 
#   ATL : 413342
#   ORD : 327943
#   DFW : 302266

airports[c(airports$iata == 'ATL'| airports$iata == 'ORD'| 
             airports$iata == 'DFW'),]
```

# Observations
With that we can begin. We note that in Q1 the month of July has the highest Arrival delay followed by December and then June. 
Likewise, in descending order, Friday, Thursday and Monday have the highest Arrival Delay. With average delays of 10.85, 10.55 and 8.28 minutes respectively. 

So, we compared the Friday of July for the top 3 airports, ATL, DFW and ORD.
## ATL, Atlanta
```{r}
#                             07/07/2005 ATL
#                               Friday
#                         DepDelay from Q1 : 10.85

toATLdep0707 <- to_depdelay(7,7, 'ATL')
i <- nrow(toATLdep0707)
toATLdep0707$cat <- rep('Departure to ATL', i)
str(toATLdep0707)

toATLarr0707 <- to_arrdelay(7,7, 'ATL')
i <- nrow(toATLarr0707)
toATLarr0707$cat <- rep('Arrival at ATL', i)
str(toATLarr0707)

toATLcombined0707 <- rbind(toATLdep0707, toATLarr0707)
toATLcombined0707

fromATLdep0707 <- from_depdelay(7,7, 'ATL')
i <- nrow(fromATLdep0707)
fromATLdep0707$cat <- rep('Departure from ATL', i)
str(fromATLdep0707)

fromATLarr0707 <- from_arrdelay(7,7, 'ATL')
i <- nrow(fromATLarr0707)
fromATLarr0707$cat <- rep('Arrival from ATL', i)
str(fromATLarr0707)

fromATLcombined0707 <- rbind(fromATLdep0707, fromATLarr0707)
fromATLcombined0707
```

```{r}
#                               Plot
#                           07/07/2005 ATL
#                               Friday

plottoATL_combine0707 <- ggplot(toATLcombined0707) +
  geom_point(aes(x = Scheduled_Time, y = Delay, color = cat),
             shape = 1, alpha = 0.5) +
  scale_color_manual(values = c('red', 'blue')) +
  geom_hline(yintercept = 10.85, color = 'red') +
  # from Q1, avg arr delay = 10.85
  geom_hline(yintercept = 0) + 
  labs(title = 'to ATL, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  scale_x_continuous(name = 'Scheduled Time', limits = c(0,2359)) +
  theme_classic() +
  theme(legend.position = 'bottom')
plottoATL_combine0707

# from ATL
plotfromATL_combine0707 <- ggplot(fromATLcombined0707) +
  geom_point(aes(x = Scheduled_Time, y = Delay, color = cat), 
             shape = 1, alpha = 0.8) +
  geom_hline(yintercept = 10.85, color = 'red') +
  # since day = 1, from Q1, avg arr delay = 10.85
  geom_hline(yintercept = 0) + 
  labs(title = 'from ATL, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  scale_x_continuous(name = 'Scheduled Time', limits = c(0,2359)) +
  theme_classic() +
  theme(legend.position = 'bottom')
plotfromATL_combine0707
```
## DFW, Dallas-Fort Worth
```{r}
#                             07/07/2005 DFW
#                               Friday
#                         DepDelay from Q1 : 10.85

toDFWdep0707 <- to_depdelay(7,7, 'DFW')
i <- nrow(toDFWdep0707)
toDFWdep0707$cat <- rep('Departure to DFW', i)
str(toDFWdep0707)

toDFWarr0707 <- to_arrdelay(7,7, 'DFW')
i <- nrow(toDFWarr0707)
toDFWarr0707$cat <- rep('Arrival at DFW', i)
str(toDFWarr0707)

toDFWcombined0707 <- rbind(toDFWdep0707, toDFWarr0707)
toDFWcombined0707

fromDFWdep0707 <- from_depdelay(7,7, 'DFW')
i <- nrow(fromDFWdep0707)
fromDFWdep0707$cat <- rep('Departure from DFW', i)
str(fromDFWdep0707)

fromDFWarr0707 <- from_arrdelay(7,7, 'DFW')
i <- nrow(fromDFWarr0707)
fromDFWarr0707$cat <- rep('Arrival from DFW', i)
str(fromDFWarr0707)

fromDFWcombined0707 <- rbind(fromDFWdep0707, fromDFWarr0707)
fromDFWcombined0707
```

```{r}
#                               Plot
#                           07/07/2005 DFW
#                               Friday

plottoDFW_combine0707 <- ggplot(toDFWcombined0707) +
  geom_point(aes(x = Scheduled_Time, y = Delay, color = cat),
             shape = 1, alpha = 0.5) +
  scale_color_manual(values = c('red', 'blue')) +
  geom_hline(yintercept = 10.85, color = 'red') +
  # from Q1, avg arr delay = 10.85
  geom_hline(yintercept = 0) + 
  labs(title = 'to DFW, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  scale_x_continuous(name = 'Scheduled Time', limits = c(0,2359)) +
  theme_classic() +
  theme(legend.position = 'bottom')
plottoDFW_combine0707

# from DFW
plotfromDFW_combine0707 <- ggplot(fromDFWcombined0707) +
  geom_point(aes(x = Scheduled_Time, y = Delay, color = cat), 
             shape = 1, alpha = 0.8) +
  geom_hline(yintercept = 10.85, color = 'red') +
  # since day = 1, from Q1, avg arr delay = 10.85
  geom_hline(yintercept = 0) + 
  labs(title = 'from DFW, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  scale_x_continuous(name = 'Scheduled Time', limits = c(0,2359)) +
  theme_classic() +
  theme(legend.position = 'bottom')
plotfromDFW_combine0707
```

## ORD, Chicago
```{r}
#                             07/07/2005 ORD
#                               Friday
#                         DepDelay from Q1 : 10.85

toORDdep0707 <- to_depdelay(7,7, 'ORD')
i <- nrow(toORDdep0707)
toORDdep0707$cat <- rep('Departure to ORD', i)
str(toORDdep0707)

toORDarr0707 <- to_arrdelay(7,7, 'ORD')
i <- nrow(toORDarr0707)
toORDarr0707$cat <- rep('Arrival at ORD', i)
str(toORDarr0707)

toORDcombined0707 <- rbind(toORDdep0707, toORDarr0707)
toORDcombined0707

fromORDdep0707 <- from_depdelay(7,7, 'ORD')
i <- nrow(fromORDdep0707)
fromORDdep0707$cat <- rep('Departure from ORD', i)
str(fromORDdep0707)

fromORDarr0707 <- from_arrdelay(7,7, 'ORD')
i <- nrow(fromORDarr0707)
fromORDarr0707$cat <- rep('Arrival from ORD', i)
str(fromORDarr0707)

fromORDcombined0707 <- rbind(fromORDdep0707, fromORDarr0707)
fromORDcombined0707
```

```{r}
#                               Plot
#                           07/07/2005 ORD
#                               Friday

plottoORD_combine0707 <- ggplot(toORDcombined0707) +
  geom_point(aes(x = Scheduled_Time, y = Delay, color = cat),
             shape = 1, alpha = 0.5) +
  scale_color_manual(values = c('red', 'blue')) +
  geom_hline(yintercept = 10.85, color = 'red') +
  # from Q1, avg arr delay = 10.85
  geom_hline(yintercept = 0) + 
  labs(title = 'to ORD, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  scale_x_continuous(name = 'Scheduled Time', limits = c(0,2359)) +
  theme_classic() +
  theme(legend.position = 'bottom')
plottoORD_combine0707

# from ORD
plotfromORD_combine0707 <- ggplot(fromORDcombined0707) +
  geom_point(aes(x = Scheduled_Time, y = Delay, color = cat), 
             shape = 1, alpha = 0.8) +
  geom_hline(yintercept = 10.85, color = 'red') +
  # since day = 1, from Q1, avg arr delay = 10.85
  geom_hline(yintercept = 0) + 
  labs(title = 'from ORD, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  scale_x_continuous(name = 'Scheduled Time', limits = c(0,2359)) +
  theme_classic() +
  theme(legend.position = 'bottom')
plotfromORD_combine0707
```

# Further Investigation
We did some testings. Our objective now is to compare if the mean avg delay of the previous period has any correlation. And to see the number of flights in each period.

We reuse the function in q1 to categorise each observation into their timebin. 
```{r}
time <- function(x){
  if (x < 100){
    return('0000-0100')
  }else if (x >= 100 & x < 200){
    return('0100-0200')
  }else if (x >= 200 & x < 300){
    return('0200-0300')
  }else if (x >= 300 & x < 400){
    return('0300-0400')
  }else if (x >= 400 & x < 500){
    return('0400-0500')
  }else if (x >= 500 & x < 600){
    return('0500-0600')
  }else if (x >= 600 & x < 700){
    return('0600-0700')
  }else if (x >= 700 & x < 800){
    return('0700-0800')
  }else if (x >= 800 & x < 900){
    return('0800-0900')
  }else if (x >= 900 & x < 1000){
    return('0900-1000')
  }else if (x >= 1000 & x < 1100){
    return('1000-1100')
  }else if (x >= 1100 & x < 1200){
    return('1100-1200')
  }else if (x >= 1200 & x < 1300){
    return('1200-1300')
  }else if (x >= 1300 & x < 1400){
    return('1300-1400')
  }else if (x >= 1400 & x < 1500){
    return('1400-1500')
  }else if (x >= 1500 & x < 1600){
    return('1500-1600')
  }else if (x >= 1600 & x < 1700){
    return('1600-1700')
  }else if (x >= 1700 & x < 1800){
    return('1700-1800')
  }else if (x >= 1800 & x < 1900){
    return('1800-1900')
  }else if (x >= 1900 & x < 2000){
    return('1900-2000')
  }else if (x >= 2000 & x < 2100){
    return('2000-2100')
  }else if (x >= 2100 & x < 2200){
    return('2100-2200')
  }else if (x >= 2200 & x < 2300){
    return('2200-2300')
  }else if (x >= 2300 & x < 2400){
    return('2300-2359')
  }else{
    return(NA)
  }
}
```

#                    Our focus will be on ATL on 07072005
## toATL

```{r}
# Creating a Timebin column and populate using the above function
toATLcombined0707$Timebin <- sapply(toATLcombined0707$Scheduled_Time, 
                                   FUN = time)
# checking the unique values
length(unique(toATLcombined0707$Timebin))
unique(toATLcombined0707$Timebin)
toATLcombined0707
```
```{r}
# to find the mean of each cat
test <- toATLcombined0707 %>%
  filter(cat == 'Arrival at ATL') %>% 
  group_by(Timebin) %>%
  summarise(count_toATL = n(),
            avg_toATL = round(mean(Delay),2)) 
test
```

## fromATL
```{r}
# Creating a Timebin column and populate using the above function
fromATLcombined0707$Timebin <- sapply(fromATLcombined0707$Scheduled_Time, 
                                    FUN = time)
# checking the unique values
length(unique(fromATLcombined0707$Timebin))
unique(fromATLcombined0707$Timebin)
 fromATLcombined0707
```
```{r}
# to find the mean of each cat
test2 <- fromATLcombined0707 %>%
  filter(cat == 'Arrival from ATL') %>% 
  group_by(Timebin) %>%
  summarise(count_fromATL = n(),
            avg_fromATL = round(mean(Delay),2))
test2
```
```{r}
# combining both test & test2 together
ttt <- test %>% 
  full_join(test2) %>% 
  arrange(Timebin)
ttt

# We rearrange the order of the time bin. Focus on flights from 0500-2359.
ttt2 <- ttt[3:21,]
ttt2
#write.csv(ttt2,'q4ATL.csv')
```
## Plots for visualisation
```{r}
# We want to combine both to and from, and count the number of flights
## to ATL
check1 <- toATLcombined0707 %>% 
  filter(cat == 'Arrival at ATL', Timebin != '0000-0100', Timebin != '0100-0200' )
## from ATL
check2 <- fromATLcombined0707 %>% 
  filter(cat == 'Arrival from ATL', Timebin != '0000-0100', Timebin != '0100-0200')

## gross
check <- rbind.data.frame(check1, check2)
check <- check %>% select(Timebin,Scheduled_Time,TailNum,Delay,cat)
check
```

```{r}
plot4 <- ggplot(check)
plot4 + 
  geom_point(aes(x=Scheduled_Time,
                 y=Delay, 
                 color = cat), alpha=0.4) + 
  geom_hline(yintercept = 60, color = 'red') +
  geom_hline(yintercept = 45, color = 'blue') +
  geom_hline(yintercept = 10.85, color = 'black') +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = 'ATL, Fri 07/07/2005', x = 'Scheduled Time', y = 'Delay') +
  theme_classic()

plot4 + 
  geom_bar(aes(x=Timebin, 
               color = cat,
               fill=cat,
               ),
           position = position_dodge2(width = 0.9)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = 'Count of Flights', x = 'Time Bin', y = 'Count')

```


### Disconnecting from db
```{r}
dbDisconnect(conn)
```
