---
title: "Q1"
author: "Gerald"
date: "2/24/2022"
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:/Users/geral/Repository/Coursework/Dataset/")
```

We first ensure our working directory is correct, and launch the libraries required.
```{r setting-wd}
getwd()
setwd("C:/Users/geral/Repository/Coursework/Dataset")
```

```{r library}
library(DBI)
library(RSQLite)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyr)
library(tidyverse)
library(RColorBrewer)
```


```{r}
display.brewer.all(colorblindFriendly = TRUE)
colours<-brewer.pal(n = 12, name = "Paired")
```

Connecting to DB Browser
```{r}
conn <- dbConnect(RSQLite::SQLite(), "dataset.db")
```

Extracting our data from 2005 to 2006 and uploading it in the global environment. 
```{r}
Y05_06 <- dbGetQuery(conn, "
                SELECT *
                FROM 'Y05_06'")
str(Y05_06)
summary(Y05_06)
head(Y05_06, n = 10)

Y05_06_narm <- na.omit(Y05_06)
Y05_06_narm
```
```{r}
# Creating Date and Day
Y05_06_narm$Date<-as.Date(with(Y05_06_narm, 
                               paste(Year, Month, DayofMonth, sep="-")),
                          "%Y-%m-%d")
Y05_06_narm$Day <- weekdays(Y05_06_narm$Date) 

Y05_06_narm <- Y05_06_narm %>% 
  filter(ArrTime<=2359,DepTime<=2359)
Y05_06_narm
# observations : 13,979,513
```
```{r}
overview <- Y05_06_narm %>% 
  group_by(Year) %>% 
  summarise(flights = n(),
            avg_arr_delay = mean(ArrDelay),
            max_arr_delay = max(ArrDelay),
            min_arr_delay = min(ArrDelay))
overview
# write.csv(overview,'overview.csv')
```

# 1. When is the best time of day, day of the week, and time of year to fly to minimise delays?

## Best Time of the Day
For best time of the day, we will be creating a barplot We separate the day into 24 periods, each period is 1 hour. Moving forward, We shall call periods as time_bin or time bin.
### Setting up the dataset 
```{r test-dataset}
test <- Y05_06_narm %>% 
  select(TailNum, ArrTime, ArrDelay, Cancelled, Diverted, Dest) %>% 
  filter(Cancelled == 0, Diverted == 0) %>% 
  arrange(ArrTime)
```

We create a function to categorise each observation into a Time bin. 
```{r}
time <- function(x){
  if (x < 100){
    return('0000-0100')
  }else if (x >= 100 & x < 200){
    return('0100-0200')
  }else if (x >= 200 & x < 300){
    return('0200-0300')
  }else if (x >= 300 & x < 400){
    return('0300-0400')
  }else if (x >= 400 & x < 500){
    return('0400-0500')
  }else if (x >= 500 & x < 600){
    return('0500-0600')
  }else if (x >= 600 & x < 700){
    return('0600-0700')
  }else if (x >= 700 & x < 800){
    return('0700-0800')
  }else if (x >= 800 & x < 900){
    return('0800-0900')
  }else if (x >= 900 & x < 1000){
    return('0900-1000')
  }else if (x >= 1000 & x < 1100){
    return('1000-1100')
  }else if (x >= 1100 & x < 1200){
    return('1100-1200')
  }else if (x >= 1200 & x < 1300){
    return('1200-1300')
  }else if (x >= 1300 & x < 1400){
    return('1300-1400')
  }else if (x >= 1400 & x < 1500){
    return('1400-1500')
  }else if (x >= 1500 & x < 1600){
    return('1500-1600')
  }else if (x >= 1600 & x < 1700){
    return('1600-1700')
  }else if (x >= 1700 & x < 1800){
    return('1700-1800')
  }else if (x >= 1800 & x < 1900){
    return('1800-1900')
  }else if (x >= 1900 & x < 2000){
    return('1900-2000')
  }else if (x >= 2000 & x < 2100){
    return('2000-2100')
  }else if (x >= 2100 & x < 2200){
    return('2100-2200')
  }else if (x >= 2200 & x < 2300){
    return('2200-2300')
  }else if (x >= 2300 & x < 2400){
    return('2300-2359')
  }else{
    return(NA)
  }
}
```

```{r}
test$Time_Bin <- sapply(test$ArrTime, FUN = time)
test
#observations : 13,979,513
```
Summarising our data for the best time of the day.
```{r}
BTOTD <- test %>% 
  group_by(Time_Bin) %>% 
  summarise(Distinct_aircraft = length(unique(TailNum)),
            No._of_flights = n(),
            Avg_Delay = round(mean(ArrDelay),2))
BTOTD
# write.csv(BTOTD,'BTOTD19.csv')
```

4) Best Time of the day plot 

```{r}
BTOTD_plot <- ggplot(BTOTD) + 
  geom_bar(aes(x = Time_Bin, y = Avg_Delay), 
           stat = 'identity', fill = colours[2], color = 'black') +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Average Arrival Delay (Intra Day)", x = " Time Bin", 
       y = " Average Delay (Mins)")
BTOTD_plot
```

## Best Day of the Week
Solving the second part of the question: "Best day of the week to travel". The following codes gives us the average departure delays. Currently Days 2,6 have the lowest departure delay, where 2 is tues & 6 is sat

```{r}
BDoWeek <- Y05_06_narm %>% 
  select(DayOfWeek, Day, TailNum, ArrDelay) %>% 
  group_by(DayOfWeek, Day) %>% 
  summarise(No_of_Flights = length(TailNum),
            Average_Delay = round(mean(ArrDelay, na.rm = TRUE),2)) 

BDoWeek
#Best day of the week 
# write.csv(BDoWeek, 'BDOW_180222.csv')
```

## Creating our plot
Creating a new column with the name for the days
```{r}
Best_day_of_the_Week <- Y05_06_narm %>%
  select(Day, DayOfWeek, ArrDelay, Cancelled, Diverted) 
Best_day_of_the_Week
```

```{r}
Best_day_of_the_Week2 <- Best_day_of_the_Week %>% 
  group_by(DayOfWeek, Day) %>%
  filter(Cancelled == 0, Diverted == 0) %>%   
  summarize(Max_Delay = max(ArrDelay, na.rm = TRUE),
            Min_Delay = min(ArrDelay, na.rm = TRUE),
            Avg_Delay = mean(ArrDelay, na.rm = TRUE),
            Q1 = quantile(ArrDelay, 0.25, na.rm = TRUE),
            Q2 = quantile(ArrDelay, 0.5,na.rm = TRUE),
            Q3 = quantile(ArrDelay, 0.75, na.rm = TRUE)) %>% 
  arrange(DayOfWeek)
Best_day_of_the_Week2
```

```{r}
#Best_day_of_the_Week2$DayOfWeek <- as.factor(Best_day_of_the_Week2$DayOfWeek)
str(Best_day_of_the_Week2)

#Best_day_of_the_Week2
```

```{r}
BDOTW_plot <- ggplot(Best_day_of_the_Week2) + 
  geom_bar(aes(x = DayOfWeek, y = Avg_Delay), 
           stat = 'identity', fill = colours[2], color = 'black') +
  theme_classic() + 
  labs(title = "Average Arrival Delay (Day)", x = "Day", 
       y = " Average Delay (Mins)") + 
  geom_text(aes(x = DayOfWeek, y = Avg_Delay, label = round(Avg_Delay,2)),
            vjust = -0.5) +
  scale_x_discrete(name = 'Day', limits = Best_day_of_the_Week2$Day)

BDOTW_plot
```

## Best Time of the Year
Solving the last part of the question: "Best time of the year to travel" 
The following codes gives us the average arrival delays. Currently Months 4,5,9 have the lowest departure delay. i.e Apr, May, Sep

```{r}
BToYear <- Y05_06_narm %>%
  mutate(Month2 = recode(Month, 
                         '1'='Jan',
                         '2'='Feb',
                         '3'='Mar',
                         '4'='Apr',
                         '5'='May',
                         '6'='Jun',
                         '7'='Jul',
                         '8'='Aug',
                         '9'='Sep',
                         '10'='Oct',
                         '11'='Nov',
                         '12'='Dec')) 
  
BToYear <- BToYear %>% 
  select(Month, Month2, TailNum, ArrDelay) %>% 
  group_by(Month, Month2) %>% 
  summarise(No_of_Flights = length(TailNum),
            Average_Delay = round(mean(ArrDelay, na.rm = TRUE),2)) 

BToYear
#Best day of the week 
#write.csv(BToYear, 'BToYear_190322.csv')
```

## Creating our plot
```{r}
Best_time_of_Year <- Y05_06_narm %>% 
  select(ArrDelay, Month, Cancelled, Diverted) %>% 
  mutate(Month2 = recode(Month, 
                         '1'='Jan',
                         '2'='Feb',
                         '3'='Mar',
                         '4'='Apr',
                         '5'='May',
                         '6'='Jun',
                         '7'='Jul',
                         '8'='Aug',
                         '9'='Sep',
                         '10'='Oct',
                         '11'='Nov',
                         '12'='Dec'))
Best_time_of_Year
```

```{r}
Best_time_of_Year2 <- Best_time_of_Year %>% 
  group_by(Month, Month2) %>% 
  filter(Cancelled == 0, Diverted == 0) %>%   
  summarize(Max_Delay = max(ArrDelay, na.rm = TRUE),
            Min_Delay = min(ArrDelay, na.rm = TRUE),
            Avg_Delay = mean(ArrDelay, na.rm = TRUE),
            Q1 = quantile(ArrDelay, 0.25, na.rm = TRUE),
            Q2 = quantile(ArrDelay, 0.5,na.rm = TRUE),
            Q3 = quantile(ArrDelay, 0.75, na.rm = TRUE)) %>% 
  arrange(Month) 
Best_time_of_Year2
```

```{r}
#Best_time_of_Year2$Month <- as.factor(Best_time_of_Year2$Month)

str(Best_time_of_Year2)
Best_time_of_Year2
```


```{r}
BTOTY_plot <- ggplot(Best_time_of_Year2) + 
  geom_bar(aes(x = Month, y = Avg_Delay), 
           stat = 'identity', fill = colours[2], color = 'black') +
  theme_classic() + 
  labs(title = "Average Arrival Delay (Month)", x = "Month", 
       y = " Average Delay (Mins)") + 
  geom_text(aes(x = Month, y = Avg_Delay, label = round(Avg_Delay,2)),
            vjust = -0.5) +
  scale_x_discrete(name = 'Day', limits = Best_time_of_Year2$Month2)
BTOTY_plot
```

### Disconnecting from db
```{r}
dbDisconnect(conn)
```