---
title: "Q2"
author: "Gerald"
date: "2/26/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:/Users/geral/Repository/Coursework/Dataset/")
```

```{r setting-wd}
getwd()
setwd("C:/Users/geral/Repository/Coursework/Dataset")
```

```{r library}
library(DBI)
library(RSQLite)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyr)
library(tidyverse)
library(RColorBrewer)
library(car)
library(broom)
```

```{r}
display.brewer.all(colorblindFriendly = TRUE)
colours<-brewer.pal(n = 12, name = "Paired")
colours
```

```{r}
# Connecting to DB Browser
conn <- dbConnect(RSQLite::SQLite(), "dataset.db")
```

# 2. Do older planes suffer more delays?

We will proceed by creating a regression model with arrival delay as the dependent variable. A dummy variable "old" will be created, and subsequently we can test the coefficient of the dummy variable to see if it is statistically significant. A significant result would imply that old planes suffer more delays. 

One concern is we do not actually have a variable that indicates whether a plane is or not, hence we will look at the mean / median year of the planes and then determining a suitable year to consider which planes are old.

## Setting up the data
Combining our dataset, Y05_06, with the plane data that shows the year of the plane. 
```{r}
Y05_6 <- dbGetQuery(conn, 
                "SELECT Y05_06.Year, Month, DayofMonth, DayOfWeek, ArrTime, DepTime, CRSArrTime, CRSDepTime, CRSElapsedTime, UniqueCarrier, Y05_06.TailNum, ArrDelay, DepDelay, Cancelled, Diverted, manufacturer, model, status, plane_data.year AS Year_Manufactured
                FROM Y05_06 JOIN plane_data ON Y05_06.TailNum=plane_data.tailnum")
str(Y05_6)
head(Y05_6)
# observations : 11,295,340 
```
```{r Carriers}
Carriers <- dbGetQuery(conn, 'SELECT * 
                      FROM Carriers')
str(Carriers)

# Observations : 1,491 
```

```{r}
Y05_6_ncarriers <- Y05_6 %>% 
  full_join(Carriers, by = c('UniqueCarrier' = 'Code')) 

Y05_6_ncarriers$Year_Manufactured <- as.integer(Y05_6_ncarriers$Year_Manufactured) 

str(Y05_6_ncarriers)

# Observations : 11,296,810 
```

## Age of the plane
Age of the plane can be found by taking Year - Year_Manufactured
```{r}
Y05_06_nage <- Y05_6_ncarriers %>% 
  filter(ArrTime <= 2359, DepTime <= 2359) %>% 
  mutate(Age = Year - Year_Manufactured)

str(Y05_06_nage)
# Observations : 11,177,955 

Y05_06narm <- na.omit(Y05_06_nage)
str(Y05_06narm)

# Observations : 10,218,859  
```

Creating a table on db browser based on the above data.

```{r}
dbRemoveTable(conn, "Q2")
dbWriteTable(conn, "Q2", Y05_06narm)
dbListFields(conn, "Q2")

rm(Carriers, Y05_06_nage, Y05_06narm, Y05_6, Y05_6_ncarriers)
```

### Q2
We can now directly extract the data required from the table in the data base, Q2.

```{r}
Y05_06narm <- dbGetQuery(conn, 'SELECT * 
                FROM Q2')
```

### To find the count of flights handled by planes according to their manufactured year

```{r}
check <- Y05_06narm %>%  
  filter(Age != 'na',Age >= '0', Age != '2005', Age != '2006',
         Age != 'NA', DepDelay != 'NA', ArrDelay != 'NA', 
         status != 'Registered to Mfr', status != 'NA')
str(check)

# Observations : 10,205,851   

summary(check)
# Year manufactured Mean : 1996 
```

```{r}
year_count <- check %>% 
  count(Year_Manufactured) %>% 
  group_by(Year_Manufactured) %>% 
  arrange()

year_count

apply(year_count['n'], MARGIN = 2, FUN = sum)
```
```{r}
barplotyearcount <- ggplot(check) +
  geom_bar(aes(x = Year_Manufactured), fill = colours[2], colour = 'black') + 
  labs(title = 'Count of Flights by Plane Manufactured Year', 
       x = 'Year Manufactured',
       y = 'Count') +
  theme_classic()
barplotyearcount
```

### Tidying up the data
Previously we found that the mean Year_Manufactured is 1996. Thus we take the mean as 1996, so planes earlier than 1996 (included) is considered old. In terms of age, planes that are older than 10 years (included) is defined as old.

We remove from dataset, observations that contains NA & 0
```{r}
A <- Y05_06narm %>% 
  select(ArrDelay, DepDelay, CRSArrTime, CRSDepTime, CRSElapsedTime, Month, DayofMonth, DayOfWeek, Age, status, manufacturer, Year_Manufactured,Cancelled, Diverted) %>% 
  filter(Age != 'na',Age >= '0', Age != '2005', Age != '2006',
         Age != 'NA', DepDelay != 'NA',
         ArrDelay != 'NA', status != 'Registered to Mfr', status != 'NA')

str(A)
A
# Observations : 10,205,851 
```
### Dummy Variable
Creating a dummy variable Old where,
      Old = 1 : planes with age >= 10 (Below 1996 (include))
      Old = 0 : planes with age < 10  (Above 1996)

```{r DVCreation}
A_dv <- A %>% 
  mutate(old = Age)

A_dv$old <- as.integer(ifelse(A_dv$old >= 10, 1 ,0))
str(A_dv)
A_dv

# Observations : 10,205,851
```
We run the following code to check how many planes are old. We found that there are more new planes than old planes. The figures are as follows: 
  Old       n
  0     6,413,027
  1     3,792,824
```{r}
old_count <- A_dv %>% 
  count(old) %>% 
  group_by(old)
old_count
```


Before we proceed with the regression, we want to check the extreme ends of the data set based on DepDelay
```{r}
check_depdelay <- quantile(A_dv$DepDelay, prob = c(0.001, 0.25, 0.5, 0.75, 0.999))
check_depdelay

summary(A_dv$DepDelay)
```
We look at planes in the 0.1 percentile and beyond the 99.9 percentile.
```{r}
# 0.1%
check1 <- A_dv %>% 
  filter(DepDelay < -17) %>% 
  arrange(DepDelay)

check1$old <- as.factor(check1$old)

str(check1)
# Observations : 7,568 

# beyond 99.9% 
check2 <- A_dv %>% 
  filter(DepDelay > 303) %>% 
  arrange()

check2$old <- as.factor(check2$old)

str(check2)
# Observations : 10,122 
```
The plots are as follows:
```{r}
# 0.1%
ggcheck1 <- ggplot(check1) +
  geom_point(aes(x = DepDelay, y = ArrDelay, colour = old)) +
  labs(title = 'Arrival vs Departure Delay (0.1 Percentile)',
       y = ' Arrival Delay', x = 'Departure Delay') + 
  theme_classic()

ggcheck1

# beyond 99.9% 
ggcheck2 <- ggplot(check2) +
  geom_point(aes(x = DepDelay, y = ArrDelay, colour = old)) +
  labs(title = 'Arrival vs Departure Delay (99.9 Percentile)',
       y = ' Arrival Delay', x = 'Departure Delay') + 
  theme_classic()

ggcheck2
```
We observe that in the 0.1 percentile chart, there are extreme values, consider the following where departure delay is less than 500 minutes while arrival delay is around 250 minutes and above. Which implies that a plane which departs early by more than 8 hours (~500 mins), can still arrive late by about 4 hours (~250mins) or more. This does not make sense hence we will remove data of planes with departure delay below -250. 

### Finalising Data set
```{r}
Final <- A_dv %>% 
  filter(DepDelay > -250)
str(Final)
Final

# Observations : 10,205,708
```
### Correlation
At first glance, we see that old has no correlation with the Arrival delay. 
```{r}
corr <- Final[c('ArrDelay','DepDelay', 'CRSArrTime', 'CRSDepTime', 'CRSElapsedTime', 'Month', 'DayofMonth', 'DayOfWeek', 'Age','Year_Manufactured', 'old')]

str(corr)
corr_table <- as.data.frame(round(cor(corr), 2))
corr_table
# write.csv(corr_table, 'corr_table.csv')
```
A plot for correlation
```{r}
#selecting a random sample of 15000 for plot
sample <- sample_n(Final, 15000)
sample_plot <- sample[c('ArrDelay','DepDelay', 'CRSArrTime', 'CRSDepTime', 'CRSElapsedTime','Month', 'DayofMonth', 'DayOfWeek', 'Age', 'old')]

checksampleplot <- plot(sample_plot)
checksampleplot
```


## Linear Regression
Performing a linear regression of Arrival delay on departure delay, CRSArrTime, CRSDepTime and old to find the significance of old planes and departure delays.
```{r old}
# We using this model
lmFinal <- lm(ArrDelay ~ DepDelay + CRSArrTime + CRSDepTime + old, data = Final)
summary(lmFinal)

tidylmFinal <- tidy(lmFinal)
df_lmFinal <- as.data.frame(tidylmFinal)
df_lmFinal
#write.csv(df_lmFinal, 'q2regsumm190322.csv')
```

```{r age_old}
lmFinal <- lm(ArrDelay ~ DepDelay + CRSArrTime + CRSDepTime + Age + old, data = Final)
summary(lmFinal)

tidylmFinal <- tidy(lmFinal)
df_lmFinal <- as.data.frame(tidylmFinal)
df_lmFinal
#write.csv(df_lmFinal, 'q2regsumm.csv')
```

```{r}
lmFinal <- lm(ArrDelay ~ DepDelay + CRSDepTime + old, data = Final)
summary(lmFinal)

tidylmFinal <- tidy(lmFinal)
df_lmFinal <- as.data.frame(tidylmFinal)
df_lmFinal
#write.csv(df_lmFinal, 'q2regsumm.csv')
```

Our test for Old gives us a t value of -0.4, which is not significant. Thus we say that Old planes are not statistically significant in causing arrival delays.

## Disconnecting from db
```{r}
dbDisconnect(conn)
```