---
title: "Q3"
author: "Gerald"
date: "2/26/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:/Users/geral/Repository/Coursework/Dataset/")
```

We first ensure our working directory is correct, and launch the libraries required.
```{r setting-wd}
getwd()
setwd("C:/Users/geral/Repository/Coursework/Dataset")
```

```{r}
library(DBI)
library(RSQLite)
library(ggplot2)
library(ggthemes)
library(zoo)
library(tidyverse)
library(reshape2)
```

```{r}
# Connecting to DB Browser
conn <- dbConnect(RSQLite::SQLite(), "dataset.db")
```

### Approach
My approach to this question is fairly lengthy. There are 3 parts to my solution. We start off by looking at the overall data set, we try to look for patterns as a whole and by the states. 
In the second part, we proceed to look at states that display negative growth, and we end off by looking at states with positive growth.

# Part 1
## Summary of Destination
```{r}
summ <-  dbGetQuery(conn, "
                     SELECT Year, Month, DayofMonth, DayofWeek, Origin, Dest, airports.iata, 
                     airports.airport, airports.city, airports.state, airports.country
                     FROM Y05_06 JOIN airports ON Y05_06.Dest=airports.iata
                     WHERE Cancelled = 0 AND Diverted = 0 AND ArrTime <= 2359 AND DepTime <= 2359")
summ$Origin <- as.factor(summ$Origin)
summ$Dest <- as.factor(summ$Dest)
summ$iata <- as.factor(summ$iata)
summ$airport <- as.factor(summ$airport)
summ$city <- as.factor(summ$city)
summ$state <- as.factor(summ$state)
summ$country <- as.factor(summ$country)
```

```{r}
str(summ)
# Observations : 13,979,513 
```

In the dataset there are destination that does not contain state / city
summ_narm contains 13,970,740 obs
```{r}
summ_narm <- na.omit(summ)
summ_narm
# Observations : 13,970,740

unique(summ_narm$state)
length(unique(summ_narm$state))
# states : 52

# just to confirm the number of observations
check <- summ %>% 
  filter(is.na(state))
str(check)
```
We found that MC Clellan-Palomar Airport and University Park, both in USA,  have missing values of state = NA. We proceed to remove observations of those two states.
Number of observations : 8,778

### List of the states
```{r}
states <- unique(summ_narm$state)
states[1]

typeof(states)
```

### Count of total number of flights in each state throughout '05 - '06
```{r}
str(summ_narm)
```

### The total flight in each state in each year
```{r}
summ_state_byyear <- summ_narm %>% 
  select(state, Year) %>% 
  filter(!is.na(state)) %>% 
  arrange(state) %>% 
  group_by(state, Year) %>% 
  count(state) 

summ_state_byyear 
summary(summ_state_byyear)

#write.csv(summ_state_byyear, "summ_state_byyear.csv") 
```

### The total flight in each state
We arrange the number of flights in descending order. 
The table is saved and used in the report.
```{r}
summ_state_total <- summ_narm %>% 
  select(state) %>% 
  filter(!is.na(state)) %>% 
  group_by(state) %>% 
  count(state) %>% 
  arrange(desc(n))

summ_state_total 


sum(summ_state_total$n)
# 13,970,740
# write.csv(summ_state_total, "summ_state_total.csv") 
```
# Top and bottom 3 states with the most flights 
The data frame shows the top and bottom 3 states, with the most and least number of flights.
```{r}
high3 <- head(summ_state_total,3)
btm3 <- tail(summ_state_total,3)

top_btm_3 <- rbind(high3,btm3)
top_btm_3
# write.csv(top_btm_3, "top_btm_3.csv")
```

## Visualisations
### Daily time series 
Daily chart not very meaningful hard to tell if there is an upward or downward trend. 

Making a new column for the date ( daily )
```{r}
daily <- summ_narm
daily$Date<-as.Date(paste(daily$Year,daily$Month,daily$DayofMonth,sep="-"),"%Y-%m-%d")
head(daily)
```

Creating a daily time series of all states
```{r}
ts_daily <- daily %>% 
  select(Date, Year, Month, DayofMonth, Origin, Dest, city, state) %>% 
  filter(!is.na(state)) %>% 
  arrange(Date) %>% 
  group_by(Date) %>% 
  count(Date)

names(ts_daily)[2] <- 'Count'
ts_daily
```

```{r}
plot_ts_daily <- ggplot(ts_daily, aes(x = Date, y = Count))
plot_ts_daily + 
  geom_line() +
  labs(title = 'Daily Count (Total)', y = 'Count', x = 'Date') +
  theme_classic()
```

### Creating a monthly time series 
Making a new column for the date ( month )
Chart not very meaningful, an apparent upward trend

```{r}
monthly <- summ_narm
monthly$Date<-as.yearmon(paste(monthly$Year,monthly$Month),"%Y %m")
head(monthly)
```

Creating a plot that shows the monthly flights for all the states
```{r}
ts_total_monthly <- monthly %>% 
  select(Date, Year, Month, Origin, Dest, city, state) %>% 
  filter(!is.na(state)) %>% 
  arrange(Date) %>% 
  group_by(Date, Year, Month) %>% 
  count(Month)
ts_total_monthly
```

### Plot for Monthly Count (Total)
```{r}
plot_ts_total_monthly <- ggplot(ts_total_monthly, aes(x = Date, y = n))
plot_ts_total_monthly +
  geom_line() +
  labs(title = 'Monthly Count (Total)', y = 'Count', x = 'Date') +
  theme_classic() 
```

### To see if there is seasonality
```{r}
ts_total_monthly_2 <- monthly %>% 
  select(Date, Year, Month, Origin, Dest, city, state) %>% 
  filter(!is.na(state)) %>% 
  arrange(Date) %>% 
  group_by(Month, Year) %>% 
  count(Month)
ts_total_monthly_2
```

```{r}
ts_total_monthly_2$Year <- as.factor(ts_total_monthly_2$Year)
ts_total_monthly_2$Month <- as.factor(ts_total_monthly_2$Month)

plot_ts_total_monthly2 <- ggplot(ts_total_monthly_2, 
                                 aes(x = Month, y = n, color = Year))
plot_ts_total_monthly2 +
  geom_point(aes(x = Month, y = n, color = Year)) +
  geom_line(aes(x = Month, y = n, color = Year, group = Year)) +
  labs(title = 'Monthly Count (Total)', y = 'Count', x = 'Month') + 
  scale_x_discrete(labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')) + 
  theme_classic() 
```

While it seems that there is no increasing trend throughout the 2 years, we can consider to look individually at each state. 

There are a total of 52 states
```{r}
ts_monthly <- monthly %>% 
  select(Date, Year, Month, Origin, Dest, city, state) %>% 
  filter(!is.na(state)) %>% 
  arrange(Date) %>% 
  group_by(Date, Year, Month, state) %>% 
  count(Date)
names(ts_monthly)[5] <- 'Count'
ts_monthly
```
The plot shows the monthly count of each state, alot of colors, hard to interpret

```{r}
plot_ts_monthly <- ggplot(ts_monthly, aes(x = Date, y = Count))
plot_ts_monthly + 
  geom_line(aes(color = state)) +
  labs(title = 'Monthly Count (By State)', y = 'Count', x = 'Date') +
  theme_classic()
```

### To Find Annual Year over Year, y-o-y
We want to compare the number of flights in 2006 against 2005, by each state. We do this by showing the year over year percentage change in flights in each state. 
```{r}
testann <- summ_narm %>% 
  select(Year, state) %>% 
  group_by(Year, state) %>% 
  count(state)
testann
```
This code shows us the Percentage change.
```{r}
testann_wide <- testann %>% 
  pivot_wider(
    id_cols = c(Year),
    names_from = state,
    values_from = n    
)
testann_wide

annyoy <- as.data.frame(round(testann_wide[2,]/testann_wide[1,]-1,5) * 100)
annyoy
```

```{r}
annyoy_long <- melt(annyoy, id.vars = 'Year')
annyoy_long <- rename(annyoy_long, state = 'variable', percentage = 'value')
annyoy_long['Year'] <- rep(2006, 52)
annyoy_long
```
Now we are able to find the states with negative and positive percentage changes.
```{r}
# total number of unique states
total <- unique(annyoy_long$state)
length(total)
```

```{r}
negstates <- subset(annyoy_long, percentage < 0)
negative <- unique(negstates$state)
negative
length(negative)             # 28 states are negative
```

```{r}
posstates <- subset(annyoy_long, percentage > 0)
positive <- unique(posstates$state)
positive
length(positive)             # 23 states are positive
```
# Part 2 : Negative Growth States
```{r}
# We select the states with negative growth listed in negative.
test <- as.data.frame(ts_monthly)
testneg <- test[test$state %in% negative,]

# We use dplyr to make it from long to wide
testneg2 <- testneg %>% 
  pivot_wider(
    id_cols = c(Date, Year, Month),
    names_from = state,
    values_from = Count
  )
testneg2

# We proceed to find the yoy changes, by comparing the months in 2005 and 2006.
testneg2_05 <- testneg2[1:12, 4:31]

testneg2_06 <- testneg2[13:24, 4:31]

pct_changeneg2 <- round((testneg2_06/testneg2_05 -1),5)*100
Year <- rep(2006,12)
Month <- seq(1,12)
Date <- as.yearmon(paste(Year,Month),"%Y %m")

# Once done, we compile them together to form a new data frame
pct_changeneg22 <-cbind(Date, pct_changeneg2)
pct_changeneg22
#write.csv(pct_changeneg22, 'pct_changeneg2.csv')
```

### Plot monthly yoy negative
```{r}
# We convert it back to long format for plotting
pct_changeneg2 <-cbind(Date,Year,Month, pct_changeneg2)
pct_changeneg2

pct_changeneg2_long <- melt(pct_changeneg2, id.vars = c('Date','Year','Month'))
str(pct_changeneg2_long)

# We tidy up the column names
pct_changeneg2_long <- pct_changeneg2_long %>% 
  select(Date, Year, Month, variable, value) %>% 
  rename(state = 'variable',
         percentage = 'value')
pct_changeneg2_long
```

```{r}
# We notice that LA stands out from the rest, hence we filter out LA
pct_changeneg2_long_noLA <- pct_changeneg2_long %>% 
  filter(state != 'LA')

pct_changeneg2_long_LA <- pct_changeneg2_long %>% 
  filter(state == 'LA')
 
nega_growth_plot <- ggplot() + 
  geom_line(data = pct_changeneg2_long_noLA,
            aes(x = Date, y = percentage, group = state),
            color = 'light blue') +
  geom_line(data = pct_changeneg2_long_LA,
            aes(x = Date, y = percentage),
            color = 'red') +
  labs(title = 'Year over Year, % (Negative Growth States)',
       y = 'Percentage', 
       x = 'Date') +
  theme(legend.position = "none") +
  theme_classic()
nega_growth_plot
```


# Part 3 : Positive Growth States
```{r}
# We select the states with positive growth listed in positive.
testpos <- test[test$state %in% positive ,]

# We use dplyr to make it from long to wide
testpos2 <- testpos %>% 
  pivot_wider(
    id_cols = c(Date, Year, Month),
    names_from = state,
    values_from = Count
  )
testpos2

# We proceed to find the yoy changes, by comparing the months in 2005 and 2006.
testpos2_05 <- testpos2[1:12, 4:26]

testpos2_06 <- testpos2[13:24, 4:26]

pct_changepos2 <- round((testpos2_06/testpos2_05 -1),5)*100
Year <- rep(2006,12)
Month <- seq(1,12)
Date <- as.yearmon(paste(Year,Month),"%Y %m")

# Once done, we compile them together to form a new data frame
pct_changepos22 <-cbind(Date, pct_changepos2)
pct_changepos22

#write.csv(pct_changepos22, 'pct_changepos2.csv')
```
### Plot monthly yoy positive
```{r}
# Likewise we change the format to long
pct_changepos2 <-cbind(Date, Year, Month, pct_changepos2)
pct_changepos2

pct_changepos2_long <- melt(pct_changepos2, id.vars = c('Date','Year','Month'))
pct_changepos2_long$Date<-as.yearmon(paste(pct_changepos2_long$Year,
                                           pct_changepos2_long$Month),"%Y %m")
str(pct_changepos2_long)

# Tidying up the column names
pct_changepos2_long <- pct_changepos2_long %>% 
  select(Date, Year, Month, variable, value) %>% 
  rename(state = 'variable',
         percentage = 'value')
pct_changepos2_long
```

```{r}
# We notice that HI stands out from the other states
pct_changepos2_long_noHI <- pct_changepos2_long %>% 
  filter(state != 'HI')

pct_changepos2_long_HI <- pct_changepos2_long %>% 
  filter(state == 'HI')


pos_growth_plot <- ggplot() + 
  geom_line(data = pct_changepos2_long_noHI,
            aes(x = Date, y = percentage, group = state),
            color = 'light blue') +
  geom_line(data = pct_changepos2_long_HI,
            aes(x = Date, y = percentage),
            color = 'red') +
  labs(title = 'Year over Year, % (Positive Growth States)',
       y = 'Percentage', 
       x = 'Date') +
  theme(legend.position = "none") +
  theme_classic()
pos_growth_plot
```

```{r}
# Disconnecting from db
dbDisconnect(conn)
```