---
title: "Coursework_setupMD.Rmd"
author: "Gerald"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "C:/Users/geral/Repository/Coursework/Dataset/")
```

```{r}
getwd()
setwd("C:/Users/geral/Repository/Coursework/Dataset")
```


## Setting up the data
In general, here are a few library that we will be utilizing throughout. 

```{r}
library(DBI)
library(RSQLite)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyr)
library(tidyverse)
library(RColorBrewer)
library(zoo)
library(car)
```

## Importing Data 
The following dataset were obtain from Havard Dataverse. The years which is have selected are 2005 and 2006. Due to the sheer amount of data, the loading time is very long. 
```{r}
airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
plane_data <- read.csv("plane_data.csv", header = TRUE)
var_des <- read.csv("variable_descriptions.csv", header = TRUE)
```

```{r}
Y2005 <- read.csv("2005.csv", header = TRUE)
```

```{r}
Y2006 <- read.csv("2006.csv", header = TRUE)
```

## Creation & Connecting to DB Browser
The following code, connects us to our database. 
```{r}
conn <- dbConnect(RSQLite::SQLite(), "dataset.db")

```

## Creation of tables 
We now procee to populate our database with the tables we have extracted. 
The tables have already been uploaded to the data base. The following codes are my initial setup process.
```{r}
#dbWriteTable(conn, "airports", airports)
#dbWriteTable(conn, "carriers", carriers)
#dbWriteTable(conn, "plane_data", plane_data)
#dbWriteTable(conn, "variable_description", var_des)

#dbWriteTable(conn, "2005", Y2005)

#dbWriteTable(conn, "2006", Y2006)

# Creation of data from '05-'06
# Now, we will combine the 2 years together and call it Y05_06.

#dbWriteTable(conn, "Y05_06", Y2005)
#dbAppendTable(conn, "Y05_06", Y2006)
```

Once we have created the new table, we check the tables in the database, and the fields within Y05_06, our combined dataset. Subsequently we tidy up our environment by removing the tables.
```{r}
dbListTables(conn)
dbListFields(conn, "Y05_06")
# rm("airports","carriers","plane_data","Y2005", "Y2006", "var_des")

```

# Overview of Data 
## Checking Sample Size in each year 

```{r}
A <- dbGetQuery(conn,
                "SELECT Year,COUNT(FlightNum) AS total_flight
                FROM 'Y05_06'
                GROUP BY Year")
A                
```

The above code allow us to check the total number of observation in each year. 
2005 : 7,140,596
2006 : 7,141,922
In total we have, 14,282,518 observations.

## Overview of delayed flights
```{r}
B <- dbGetQuery(conn, 
                "SELECT Year, 
                COUNT(DISTINCT TailNum) AS distinct_aircraft,
                COUNT(FlightNum) AS flights,
                AVG(ArrDelay) AS avg_arr_delay,
                MAX(ArrDelay) AS max_arr_delay,
                MIN(ArrDelay) AS min_arr_delay
                FROM 'Y05_06'
                WHERE Cancelled = 0 AND Diverted = 0
                GROUP BY Year")
B
```

```{r}
C <- A %>% 
  inner_join(B, by = "Year")
C
```

After running the above 2 code chunks, we obtain a summary of the flights in 2005 and 2006. We filtered out flights that were cancelled or diverted to prevent and distortion of our data.

Excluding cancelled and diverted flights: 
Total flights 13,996,640
2005 : 6,992,838
2006 : 7,003,802

# Extracting data from '05 to '06 and finding the structure of the data
We proceed to extract the Y05_06 dataset into our global environment, for easy reference.
```{r}
Y05_06 <- dbGetQuery(conn, "
                SELECT *
                FROM 'Y05_06'")
str(Y05_06)
summary(Y05_06)
head(Y05_06, n = 10)

```

Total observations : 14,282,518

### Checking if the dataset consist of only the USA
We filter out all the destination, dest, listed in Y05_06. To cross check with the list of airports. This can be done by creating a data frame 'check' and then parsing in our dest and inner join with the airports.
```{r}
dest <- dbGetQuery(conn,"
                   SELECT Dest
                   FROM Y05_06")
```

```{r}
airports <- dbGetQuery(conn, "
                       SELECT iata, airport, state, country
                       FROM airports")
```

```{r}
check <- dest %>% 
  inner_join(airports, by = c("Dest" = "iata"))

check$Dest <- as.factor(check$Dest)
check$airport <- as.factor(check$airport)
check$state <- as.factor(check$state)
check$country <- as.factor(check$country)

str(check)
```
```{r}
unique(check$Dest)
unique(check$airport)
unique(check$state)
unique(check$country)
```

With that we conclude the following results: 
Total Airports: 301
Total states : 52
Total Country : 1

## To confirm the total observation less na 
```{r}
check <- dbGetQuery(conn,
                "SELECT *
                FROM 'Y05_06'")
str(check)
# Observations : 14,282,518 

check_narm <- na.omit(check) 
str(check_narm)
# Observations : 13,996,640 

# Check year 
count <- check_narm %>% 
  count(Year) %>% 
  group_by(Year) 

names(count)[names(count) == 'n'] <- 'count'
count
```
When we remove the observations that contain NA, we have a total of 13,996,640 observations.

To ensure that the data set does not contains any cancelled and/or diverted flights 
```{r}
check_narm1 <- check_narm %>% 
  filter(Cancelled == 0, Diverted == 0)
check_narm1
```

# Disconnecting from db
```{r}
dbDisconnect(conn)
```